---
title: "CM 2605: Simulation and Modeling Techniques"
subtitle: "CW Part 01 — Simulation and Stochastic Analysis of Hospital Emergency Resources"
author: "J. A. N. Vishara Jayalath | RGU: 2425578 | IIT: 20240184"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Simulation and Stochastic Analysis of Hospital Emergency Resources

# Introduction

This report presents a discrete-event simulation of the Emergency Room (ER) triage process at a hospital experiencing overcrowding during its weekend shifts. The simulation models patient arrivals and service provided using appropriate stochastic distributions, and evaluates whether the current staffing level of 3 triage doctors meets the hospital board's mandated threshold of an average patient waiting time not exceeding 10 minutes.

A second scenario is also simulated with 4 triage doctors to assess and compare whether the additional resource resolves the overcrowding issue.

---

# Model Parameters

The system is modelled as an **M/M/c queue**, where:

- Arrivals follow a **Poisson process** with rate **$\lambda = 12$** patients/hour ($= 0.2$ patients/minute)
- Service times follow an **Exponential distribution** with mean $\mu = 12$ minutes per patient per doctor
- Queue discipline is **First-Come, First-Served (FCFS)**
- Number of servers $c = 3$ (Scenario 1) and $c = 4$ (Scenario 2)
- We simulate $n = 500$ patients

The server utilisation factor is:

$$\rho = \frac{\lambda}{c \cdot \mu^{-1}} = \frac{12 \text{ patients/hr}}{c \times 5 \text{ patients/hr/doctor}}$$
For $c = 3$: $\rho = \frac{12}{15} = 0.80$

For $c = 4$: $\rho = \frac{12}{20} = 0.60$

---

# Simulation Algorithm

## Helper Function: Multi-Server Queue Simulation

```{r}

# Multi- Server Queue Simulation (M/M/c)

# Arguments:
#   n_patients: number of patients to simulate
#   arrival_rate: average patients per minute (lambda)
#   service_rate: average service time in minutes per patient (1/mu)
#   seed: random seed for reproducibility of results

simulate_er <- function(n_patients = 500, 
                        arrival_rate = 12/60, # patients per minute
                        service_mean = 12, # average minutes per patient
                        n_doctors = 3,
                        seed = 42){
  set.seed(seed)
  
  
  
  # Step 1: Generate inter-arrival times (Exponential from Poisson process)
  # If arrivals are Poisson (lambda), inter-arrival gaps are Exp(lambda)
  inter_arrivals <- rexp(n = n_patients, rate = arrival_rate)
  
  # Step 2: Compute cumulative arrival times
  arrival_times <- cumsum(inter_arrivals)
  
  # Step 3: Generate service durations for each patient 
  service_durations <- rexp(n = n_patients, rate = 1 / service_mean)
  
  # Step 4: Track when each doctor becomes free
  # Initialize all doctors as free from time 0
  doctor_free_at <- rep(0,n_doctors)
  
  # Step 5: Pre-allocate result vectors
  service_start <- numeric(n_patients)
  service_end <- numeric (n_patients)
  wait_time <- numeric (n_patients)
  system_time <- numeric (n_patients)
  
  # Step 6: Process each patient sequentially (FCFS)
  
  for (i in 1:n_patients){
    
    # Find the doctor who becomes free earliest
    earliest_doctor <- which.min(doctor_free_at)
    doctor_available_at <- doctor_free_at[earliest_doctor]
    
    # Patient can only start service when BOTH the patient has arrived 
    # AND a doctor is available - whichever is later
    service_start[i] <- max(arrival_times[i], doctor_available_at)
    
    # Service ends after their service duration
    service_end[i] <- service_start[i] + service_durations[i]
    
    # Update that doctor's next free time
    doctor_free_at[earliest_doctor] <- service_end[i]
    
    # Compute waiting time (time spent in queue before being seen)
    wait_time[i] <- service_start[i] - arrival_times[i]
    
    # Compute total system time (wait + service)
    system_time[i] <- service_end[i] - arrival_times[i]
  }
  
  # Step 7: Return results as a data frame
  
  results <- data.frame(
    Patient = 1:n_patients,
    Arrival_Time = round(arrival_times, 4),
    Service_Start = round(service_start, 4),
    Service_End = round(service_end, 4),
    Service_Time = round(service_durations, 4),
    Wait_Time = round(wait_time, 4),
    System_Time = round(system_time, 4)
  )
  
  return(results)
}
```

---

# Scenario 1: Current System (3 Triage Doctors)

```{r}

# Run Scenario 1
sim1 <- simulate_er(n_patients = 500,
                    arrival_rate = 12/60,
                    service_mean = 12,
                    n_doctors = 3, 
                    seed = 42)


# Display first 10 rows for inspection

knitr:: kable(head(sim1, 10),
              caption = "Scenario 1 - First 10 Patients (3 Doctors)",
              digits = 2)
```

---

## Performance Metrics - Scenario 1

```{r}

# Key performance metrics
avg_wait_1 <- mean(sim1$Wait_Time)
avg_system_1 <- mean(sim1$System_Time)
max_wait_1 <- max(sim1$Wait_Time)
pct_exceed_1 <- mean(sim1$Wait_Time > 10) * 100

# Average Queue Length (Little's Law): L_q = Lambda * W_q
lambda_per_min <- 12/60
avg_queue_1 <- lambda_per_min * avg_wait_1

cat("  SCENARIO 1: 3 Triage Doctors\n")

cat("========================================\n")

cat(sprintf("  Average Waiting Time : %.2f minutes\n", avg_wait_1))

cat(sprintf("  Average System Time  : %.2f minutes\n", avg_system_1))

cat(sprintf("  Maximum Waiting Time : %.2f minutes\n", max_wait_1))

cat(sprintf("  Average Queue Length : %.2f patients\n", avg_queue_1))

cat(sprintf("  %% Patients Waiting >10 min : %.1f%%\n", pct_exceed_1))

cat(sprintf("  Threshold Met (<=10 min) : %s\n",
ifelse(avg_wait_1 <= 10, "YES", "NO")))

cat("========================================\n")
```
The simulation shows an average waiting time of 6.84 minutes, which is below the 10-minute threshold, indicating that the system does indeed meet the requirement on average.

However, performance is not consistent as around 24.6% of patients wait longer than 10 minutes. The maximum waiting time reaches 52.06 minutes as well, highlighting significant variability.

The average system time of 19.45 minutes and queue length of 1.37 patients suggest moderate congestion especially during peak periods.

Overall, while the system meets the target on average, it operates close to capacity and may not reliably provide acceptable service for all patients.

---

# Scenario 2: Additional Doctor (4 Triage Doctors)

```{r}

# Run Scenario 2 - set seed for fair comparison 

sim2 <- simulate_er(n_patients = 500,
                    arrival_rate = 12/60,
                    service_mean = 12,
                    n_doctors = 4,
                    seed = 42)

# Display first 10 rows

knitr::kable(head(sim2, 10),
             caption = "Scenario 2 - First 10 Patients (4 Doctors)",
             digits = 2)
```

---

## Performance Metrics - Scenario 2

```{r}

avg_wait_2   <- mean(sim2$Wait_Time)
avg_system_2 <- mean(sim2$System_Time)
max_wait_2   <- max(sim2$Wait_Time)
pct_exceed_2 <- mean(sim2$Wait_Time > 10) * 100
avg_queue_2  <- lambda_per_min * avg_wait_2


cat("  SCENARIO 2: 4 Triage Doctors\n")

cat("========================================\n")

cat(sprintf("  Average Waiting Time  : %.2f minutes\n", avg_wait_2))

cat(sprintf("  Average System Time   : %.2f minutes\n", avg_system_2))

cat(sprintf("  Maximum Waiting Time  : %.2f minutes\n", max_wait_2))

cat(sprintf("  Average Queue Length  : %.2f patients\n", avg_queue_2))

cat(sprintf("  %% Patients Waiting >10 min : %.1f%%\n", pct_exceed_2))

cat(sprintf("  Threshold Met (<=10 min)   : %s\n",
            ifelse(avg_wait_2 <= 10, "YES", "NO")))

cat("========================================\n")
```





































